Bayesian regression facilitates quantitative modelling of cell metabolism

### 2023-10-31 paper/main.qmd
Update paper

Posterior predictive checking indicated that our model achieved a good fit to
the simulated reaction and metabolite concentration measurements, as shown by
the graphs in the top row of figure @fig-posterior.

Analysis of the posterior distributions for kinetic parameters indicated that
these are highly correlated. The marginal posterior distributions for most
kinetic parameters did not shrink significantly compared with the corresponding
marginal prior distributions, even though these parameters’ joint posterior
distribution contained enough information to make accurate out of sample
predictions. In some cases, there were two-dimensional correlations such as
the one shown in the bottom left of figure @fig-posterior; in this case the
marginal distribution of the two parameters is roughly banana shaped. More
commonly, however, two-dimensional pair plots were insufficient to reveal the
underlying correlation structure, as seen in the bottom-right plot in figure
@fig-posterior. This does not mean that the parameters were uncorrelated, but
rather that the correlations involve more than two parameters.

Overall, our results show that Maud can fit a realistic pathway-sized dataset.
This was achieved without fixing the marginal values of kinetic parameters: the
information required to make good predictions was contained in the correlation
structure of the joint posterior distribution. This finding is consistent with
previous analyses of biological systems that found they are “sloppy”, that is,
sensitive to parameter combinations rather than marginal parameter values, with
important combinations, scales and regions of sensitivity being difficult to
ascertain in advance [@gutenkunst_2007; @poirier_revising_1998].

@@

The difference between the Laplace approximation output and the true posterior
distribution manifests not only in the parameter space, but also in the
measurement space. Figure @fig-laplace frame B compares the 5%-95% interval
for flux measurement log likelihoods in the true posterior with the Laplace
approximation; lower log likelihood values indicate that the modelled and
measured values are further away. The graph shows that the Laplace approximation
yielded significantly worse predictions than the true posterior, even for the
training data.

To further explore why this is the case we compared samples from the true
posterior and the Laplace approximation for the pairwise marginal distributions
of two Michaelis-Menten constants $K_{m}^{A,r1}$ and $K_{cat}^{r1}$: see the
bottom right cell of figure @fig-laplace. This comparison demonstrates that
the Laplace method is not able to capture the correct relationships between
parameters’ distributions.

>>>>

Posterior predictive checking indicated that our model achieved a good fit to
the simulated reaction and metabolite concentration measurements, as shown by
the graphs in the top row of @fig-posterior.

Analysis of the posterior distributions for kinetic parameters indicated that
these are highly correlated. The marginal posterior distributions for most
kinetic parameters did not shrink significantly compared with the corresponding
marginal prior distributions, even though these parameters’ joint posterior
distribution contained enough information to make accurate out of sample
predictions. In some cases, there were two-dimensional correlations such as
the one shown in the bottom left of @fig-posterior; in this case the
marginal distribution of the two parameters is roughly banana shaped. More
commonly, however, two-dimensional pair plots were insufficient to reveal the
underlying correlation structure, as seen in the bottom-right plot in figure
@fig-posterior. This does not mean that the parameters were uncorrelated, but
rather that the correlations involve more than two parameters.

These results show that Maud can achieve an adequate fit to a realistic
pathway-sized dataset. This was achieved without fixing the marginal values
of kinetic parameters: the information required to make good predictions was
contained in the correlation structure of the joint posterior distribution.
This finding is consistent with previous analyses of biological systems that
found they are “sloppy”, that is, sensitive to parameter combinations rather
than marginal parameter values, with important combinations, scales and regions
of sensitivity being difficult to ascertain in advance [@gutenkunst_2007;
@poirier_revising_1998].

@@

The difference between the Laplace approximation output and the true posterior
distribution manifests not only in the parameter space, but also in the
measurement space. The upper right (training data) and lower left (validation
data) plots in @fig-laplace show 5%-95% marginal flux measurement log likelihood
intervals for both methods; lower log likelihood values indicate that the
modelled and measured values are further away. The graph shows that the Laplace
approximation yielded significantly worse predictions than the true posterior,
even for the training data.

To further explore why this is the case we compared samples from the true
posterior and the Laplace approximation for the pairwise marginal distributions
of two Michaelis-Menten constants $K_{m}^{A,r1}$ and $K_{cat}^{r1}$: see the
bottom right cell of @fig-laplace. This comparison demonstrates that
the Laplace method is not able to capture the correct relationships between
parameters’ distributions.

### 2023-10-31 paper/main.tex
Update paper

\title[]{Bayesian regression facilitates quantitative modelling of cell
metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Our paper reviews the existing options in this
area, presents a case study illustrating how Maud can be used to analyse
a metabolic network and explains the biological, statistical and
computational design decisions underpinning Maud.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, breakable]}{\end{tcolorbox}}\fi

@@

Posterior predictive checking indicated that our model achieved a good
fit to the simulated reaction and metabolite concentration measurements,
as shown by the graphs in the top row of figure
Figure~\ref{fig-posterior}.

Analysis of the posterior distributions for kinetic parameters indicated
that these are highly correlated. The marginal posterior distributions
for most kinetic parameters did not shrink significantly compared with
the corresponding marginal prior distributions, even though these
parameters' joint posterior distribution contained enough information to
make accurate out of sample predictions. In some cases, there were
two-dimensional correlations such as the one shown in the bottom left of
figure Figure~\ref{fig-posterior}; in this case the marginal
distribution of the two parameters is roughly banana shaped. More
commonly, however, two-dimensional pair plots were insufficient to
reveal the underlying correlation structure, as seen in the bottom-right
plot in figure Figure~\ref{fig-posterior}. This does not mean that the
parameters were uncorrelated, but rather that the correlations involve
more than two parameters.

Overall, our results show that Maud can fit a realistic pathway-sized
dataset. This was achieved without fixing the marginal values of kinetic
parameters: the information required to make good predictions was
contained in the correlation structure of the joint posterior
distribution. This finding is consistent with previous analyses of
biological systems that found they are ``sloppy'', that is, sensitive to
parameter combinations rather than marginal parameter values, with
important combinations, scales and regions of sensitivity being
difficult to ascertain in advance
\citep{gutenkunst_2007, poirier_revising_1998}.

@@

The difference between the Laplace approximation output and the true
posterior distribution manifests not only in the parameter space, but
also in the measurement space. Figure Figure~\ref{fig-laplace} frame B
compares the 5\%-95\% interval for flux measurement log likelihoods in
the true posterior with the Laplace approximation; lower log likelihood
values indicate that the modelled and measured values are further away.
The graph shows that the Laplace approximation yielded significantly
worse predictions than the true posterior, even for the training data.

To further explore why this is the case we compared samples from the
true posterior and the Laplace approximation for the pairwise marginal
distributions of two Michaelis-Menten constants \(K_{m}^{A,r1}\) and
\(K_{cat}^{r1}\): see the bottom right cell of figure
Figure~\ref{fig-laplace}. This comparison demonstrates that the Laplace
method is not able to capture the correct relationships between
parameters' distributions.

>>>>

\title[]{Bayesian regression facilitates quantitative modelling of cell
metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Our paper reviews the existing options in this
area, presents a case study illustrating how Maud can be used to analyse
a metabolic network and explains the biological, statistical and
computational design decisions underpinning Maud.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, boxrule=0pt, interior hidden, breakable, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

Posterior predictive checking indicated that our model achieved a good
fit to the simulated reaction and metabolite concentration measurements,
as shown by the graphs in the top row of Figure~\ref{fig-posterior}.

Analysis of the posterior distributions for kinetic parameters indicated
that these are highly correlated. The marginal posterior distributions
for most kinetic parameters did not shrink significantly compared with
the corresponding marginal prior distributions, even though these
parameters' joint posterior distribution contained enough information to
make accurate out of sample predictions. In some cases, there were
two-dimensional correlations such as the one shown in the bottom left of
Figure~\ref{fig-posterior}; in this case the marginal distribution of
the two parameters is roughly banana shaped. More commonly, however,
two-dimensional pair plots were insufficient to reveal the underlying
correlation structure, as seen in the bottom-right plot in figure
Figure~\ref{fig-posterior}. This does not mean that the parameters were
uncorrelated, but rather that the correlations involve more than two
parameters.

These results show that Maud can achieve an adequate fit to a realistic
pathway-sized dataset. This was achieved without fixing the marginal
values of kinetic parameters: the information required to make good
predictions was contained in the correlation structure of the joint
posterior distribution. This finding is consistent with previous
analyses of biological systems that found they are ``sloppy'', that is,
sensitive to parameter combinations rather than marginal parameter
values, with important combinations, scales and regions of sensitivity
being difficult to ascertain in advance
\citep{gutenkunst_2007, poirier_revising_1998}.

@@

The difference between the Laplace approximation output and the true
posterior distribution manifests not only in the parameter space, but
also in the measurement space. The upper right (training data) and lower
left (validation data) plots in Figure~\ref{fig-laplace} show 5\%-95\%
marginal flux measurement log likelihood intervals for both methods;
lower log likelihood values indicate that the modelled and measured
values are further away. The graph shows that the Laplace approximation
yielded significantly worse predictions than the true posterior, even
for the training data.

To further explore why this is the case we compared samples from the
true posterior and the Laplace approximation for the pairwise marginal
distributions of two Michaelis-Menten constants \(K_{m}^{A,r1}\) and
\(K_{cat}^{r1}\): see the bottom right cell of Figure~\ref{fig-laplace}.
This comparison demonstrates that the Laplace method is not able to
capture the correct relationships between parameters' distributions.
